---
title: "Poisson Process Simulation"
output:
  pdf_document: default
  html_notebook: default
---

## Example 1
Let's set up a basic simulation for the PP.  We can run it for a fixed time or until a fixed number of events have occurred.  For multiple runs, a fixed time might more sense.  Let's set up a function to do this.
```{r}
# Fixed number of events
sim_PP_event<-function(k,lambda){
  T<-rexp(n=k,rate=lambda)
  return(cumsum(T))
}
```
Now lets try a few examples.
```{r}
k<-20; lambda<-1/2
N1<-sim_PP_event(k,lambda)
plot(stepfun(N1,0:k)) #needs starting value included
```



## Example 2
Now lets generate the process multiple times, but for a fixed time interval.  This is a little bit harder and there are at least two ways you could do it.  The first is to still only keep track of the arrival times, while second uses a grid.

### Approach 1
```{r}
# Fixed interval
sim_PP_event<-function(n,T,lambda,KMax){
  S<-vector(mode="list",length=n)
  for(i in 1:n){
    T_tmp<-rexp(n=KMax,rate=lambda)
    S_tmp<-cumsum(T_tmp)
    #if(S_tmp[1]>T){S[[i]]=NULL
    #}else{
    S[[i]] = S_tmp[S_tmp<=T]#} 
  }
  return(S)
}
```

Let's try it out.  
```{r}
set.seed(515)
n<-20; lambda<-2; T<-5; KMax<-100
N<-sim_PP_event(n,T,lambda,KMax)
NT<-unlist(lapply(N,FUN="length"))
plot(stepfun(N[[1]],0:NT[1]),ylim=c(0,max(NT)),xlim=c(0,T)) 
for(i in 2:n){
  plot(stepfun(N[[i]],0:NT[i]),add=TRUE,col=i) 
}
```
Note that this code won't work quite right if there are runs with no events.  You can fix this, but it is a bit annoying.


#### Approach 2
This isn't as elegent in terms of saving memory, but it is a bit simpler.  We will just fix a grid to generate the processes on.
```{r}
T<-1
M<-100 # number of grid points
lambda<-5
n<-1000

# Can include default for M
sim_PP_grid<-function(n,T,lambda,M=100){
  t<-seq(0,T,length=M)
  S_mat<-matrix(0,nrow=n,ncol=M)
  for(i in 1:n){
    Ntmp<-rpois(1,lambda=lambda)
    if(Ntmp>0){
      Mtmp<-sample(1:M,size=Ntmp)
      S_mat[i,Mtmp]<-1
    }
  }
  N<-t(apply(S_mat,1,FUN="cumsum"))
  return(list(t=t,N=N))
}

PP<-sim_PP_grid(n,T,lambda)
matplot(PP$t,t(PP$N),type="l")
```
Not absolutely perfect, but not bad.  Also a lot easier.

```{r}
# Lets look at number of points that fall between 0 and T/2
N12<-PP$N[,M/2]
barplot(summary(as.factor(N12))/n,ylim=c(0,0.25))
barplot(dpois(0:max(N12),lambda=lambda*T/2),names.arg=0:max(N12),ylim=c(0,0.25))
```










