---
title: "R Notebook"
output:
  pdf_document: default
  html_document:
    df_print: paged
---



To approximate the transition probability function we have to work in finite states, though infinite states can be reasonably approximated in some cases.  Let's set up the basics of birth/death process, which is basically any MP that only goes up/down one unit when it transitions.
```{r}
# Largest state
p<- 4

# Set up transition rates vectors
lambda<-numeric(p+1)
lambda[p+1] =0
mu<-numeric(p+1)
mu[1] = 0

# Choose the rates
lambda[-(p+1)] = 2
mu[-1] = 2

# Holding rates
v = lambda + mu

# Transition probabilities
P<-matrix(0,nrow=(p+1),ncol=(p+1))
spts<-cbind(1:p,2:(p+1))
P[spts] = lambda[-(p+1)]/v[-(p+1)]
spts<-cbind(2:(p+1),1:(p)) 
P[spts] = mu[-1]/v[-1]
P
```

We now set up the $\bf R$ matrix.
```{r}
R<-diag(v)%*%P
diag(R)<- -v
R
```
Using this matrix, we can approximate the transition probability function.
```{r}
#install.packages("Matrix")
library(expm)
t<-1
n<-10000
I<-diag(1,p+1,p+1)
Pt<-(I + R*t/n) %^% n
Pt
```
Let's see how thse change with $n$
```{r}
t<-1
i<-0
j<-p
n_all<-1000
pij<-numeric(n_all)
for(n in 1:n_all){
  Pt<-(I + R*t/n) %^% n
  pij[n] = Pt[i+1,j+1]
}
plot(1:n_all,pij,type="l")
```
Finally, let's verify this with MC
```{r}
n<-100
Pt<-(I + R*t/n) %^% n

i<-0
t<-1
reps<-100000
Xt<-numeric(reps)
for(r in 1:reps){
  t_c<-0
  x_c<-i
  while(t_c<t){
    v_c<-v[x_c+1]
    t_c<-t_c+rexp(1,rate=v_c)
    if(x_c!=p){pr<-P[x_c+1,x_c+2]
    }else{pr<-0}
    U<-rbinom(1,1,p=pr)
    if(U==1){x_c = x_c+1
    }else{x_c = x_c-1}
    if(t_c<t){Xt[r] = x_c}
  }
}
summary(as.factor(Xt))/reps
Pt[1,]
```


We can also use this function to simulate on a grid.  Choose an evenly spaced set of points which are $\delta t$ apart, then compute $P_{ij(\delta t)}$ and use it to simulate a discrete time chain.  
```{r}
T<-2
m<-100
times<-seq(0,T,length=m)
delta<-T/(m-1)

n<-100
Pdel<-(I + R*delta/n) %^% n

Xt<-numeric(m)
Xt[1]<-1

for(i in 2:m){
  Pr<-Pdel[Xt[i-1]+1,]
  Xt[i]<-sample(0:p,1,prob=Pr)
}
plot(times,Xt,type="l")
```

Lastly, if you want to simulate with an unbounded state space, you can approximate it with a large bounded one.  Notice that after two time units, we hit the edge of the state space.  So we can increase this.

```{r}
p<-20
lambda<-numeric(p+1)
lambda[p+1] =0
mu<-numeric(p+1)
mu[1] = 0

lambda[-(p+1)] = 2
mu[-1] = 2
v = lambda + mu

P<-matrix(0,nrow=(p+1),ncol=(p+1))
spts<-cbind(1:p,2:(p+1))
P[spts] = lambda[-(p+1)]/v[-(p+1)]
spts<-cbind(2:(p+1),1:(p)) 
P[spts] = mu[-1]/v[-1]


R<-diag(v)%*%P
diag(R)<- -v
I<-diag(1,p+1,p+1)
```

```{r}
reps<-100
T<-2
m<-100
times<-seq(0,T,length=m)
delta<-T/(m-1)

n<-100
Pdel<-(I + R*delta/n) %^% n

X_all<-matrix(nrow=reps,ncol=m)
for(r in 1:reps){
  Xt<-numeric(m)
  Xt[1]<-1
  for(i in 2:m){
    Pr<-Pdel[Xt[i-1]+1,]
    Xt[i]<-sample(0:p,1,prob=Pr)
  }
  X_all[r,] = Xt
}
matplot(times,t(X_all),type="l")
```












